\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-17 doc/mlsys_paper/paper.tex
remove some redundancy in the description of various optimizers

Initially, we implemented these functions as simply as possible and ran them
without any tuning, which reflects how a normal user might interact with the
software.  These results are seen in the first row of
Table~\ref{tab:rosenbrock_results}.  Only Julia and {\tt
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

%% CS: i'm not convinced that this conjecture is barking up the right tree;
%% CS: SciPy is in Python, which is by default an interpreted language.
%% CS: the slowdown many have nothing to do with inlining and simply due to
%% CS: Python being slow in general.
%% CS: it's possible to use PyPy or Numba to speed things up,
%% CS: in order to provide a "fairer" comparison against compiled languages.
%% CS: Octave also has an optional JIT compiler, which last time I checked
%% CS: needs to be explicitly enabled.
%% 
%% CS: i would suggest to stay away from the "we do inlining" argument,
%% CS: as it can be blown up too easily.
%% CS: furthermore, I would suggest to mention PyPy and/or Numba and Octave JIT
%% CS: (in order to show that we've done our homework),
%% CS: and then state something along the lines of
%% CS: "these approaches are not being used as we're comparing out-of-the-box performance"

% RC: agree, hopefully the paragraph below pushes this in more of a right
% direction.

However, both Python and GNU Octave have routes for acceleration, such as
numba~\cite{TODO} and MEX bindings for Octave/MATLAB.  Therefore, we hand-optimized the
Rosenbrock implementation using numba and ran the Octave implementation with the
JIT.  For the numba implementation, the underlying {\tt anneal.anneal()}
function had to be significantly modified.  These did produce some acceleration,
as can be seen in the second row of Table~\ref{tab:rosenbrock_results}.  For
Octave, a MEX binding did not produce a noticeable difference.

Next, we consider the linear regression example described in Sec.~\ref{sec:linreg_example}.
For this task we use the first-order L-BFGS optimizer~\cite{zhu1997algorithm}.
For {\tt ensmallen} we have 2 versions:
(i)~with only \texttt{\small EvaluateWithGradient()},
and
(ii)~with \texttt{\small Evaluate()} and \texttt{\small Gradient()}.
The code for these functions is as shown earlier.
For Julia we have the options of using manually defined objective and gradient functions,
or the gradient function can be automatically computed by 
{\tt Calculus.jl}
({\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl})
or {\tt ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016}.
Overall we compare across the following:

\vspace*{-0.3em}
\begin{itemize}
\renewcommand{\itemsep}{-0.5ex}
  \item {\tt ensmallen}-1: only \texttt{\small EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both \texttt{\small Evaluate()} and \texttt{\small Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{itemize}
\vspace*{-0.3em}

>>>>

Initially, we implemented these functions as simply as possible and ran them
without any tuning. This reflects how a typical user might interact with a
given framework.
The results are shown in the first row of
Table~\ref{tab:rosenbrock_results}.  Only Julia and {\tt
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

However, both Python and Octave have routes for acceleration,
such as Numba~\cite{TODO}, MEX bindings and JIT compilation.
We hand-optimized the Rosenbrock implementation using Numba,
which required significant modification of the 
underlying \texttt{\small anneal.anneal()} function.
We ran the Octave implementation with JIT enabled.
These techniques did produce some speedup,
as shown in the second row of Table~\ref{tab:rosenbrock_results}.
For Octave, a MEX binding did not produce a noticeable difference.

Next, we consider the linear regression example described in Sec.~\ref{sec:linreg_example}.
For this task we use the first-order L-BFGS optimizer~\cite{zhu1997algorithm}.
We compare across the following:

\vspace*{-0.3em}
\begin{enumerate}[{~~~$\bullet$}]
\renewcommand{\itemsep}{-0.5ex}
  \item {\tt ensmallen}-1: only \texttt{\small EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both \texttt{\small Evaluate()} and \texttt{\small Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} (\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl}) for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{enumerate}
\vspace*{-0.3em}

% For {\tt ensmallen} we have 2 versions:
% (i)~with only \texttt{\small EvaluateWithGradient()},
% and
% (ii)~with \texttt{\small Evaluate()} and \texttt{\small Gradient()}.
% The code for these functions is as shown earlier.
% For Julia we have the options of using manually defined objective and gradient functions,
% or the gradient function can be automatically computed by 
% {\tt Calculus.jl}
% (\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl})
% or {\tt ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016}.
